---
title: "statistical analysis of chr7-16-19"
author: "ilbilgeulukoy"
date: "2025-07-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# reading the files of chr7-16-19

```{r}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)  
library(dplyr)
library(janitor)
chr7 <- read_tsv("~/Downloads/chr7.tsv", col_types = cols(.default = "c")) %>% clean_names()
chr16 <- read_tsv("~/Downloads/chr16.tsv", col_types = cols(.default = "c")) %>% clean_names()
chr19 <- read_tsv("~/Downloads/chr19.tsv", col_types = cols(.default = "c")) %>% clean_names()

```

#selection of all mRNAs in chromosomes
```{r}
knitr::opts_chunk$set(echo = TRUE)

chr7_lengths <- chr7 %>%
  filter(transcript_type == "mRNA") %>%
  mutate(transcript_length = as.numeric(stop) - as.numeric(start),
         chr = "chr7") %>%
  select(chr, transcript_length)

chr16_lengths <- chr16 %>%
  filter(transcript_type == "mRNA") %>%
  mutate(transcript_length = as.numeric(stop) - as.numeric(start),
         chr = "chr16") %>%
  select(chr, transcript_length)

chr19_lengths <- chr19 %>%
  filter(transcript_type == "mRNA") %>%
  mutate(transcript_length = as.numeric(stop) - as.numeric(start),
         chr = "chr19") %>%
  select(chr, transcript_length) 
  
  combined_lengths <- bind_rows(chr7_lengths, chr16_lengths, chr19_lengths) 
  
```


#selection of all mRNAs<5000bp
```{r}
chr7_lengths <- chr7 %>%
  filter(transcript_type == "mRNA") %>%
  mutate(transcript_length = as.numeric(stop) - as.numeric(start),
         chr = "chr7") %>%
  filter(transcript_length < 5000, !is.na(transcript_length)) %>% 
  select(chr, transcript_length)

chr16_lengths <- chr16 %>%
  filter(transcript_type == "mRNA") %>%
  mutate(transcript_length = as.numeric(stop) - as.numeric(start),
         chr = "chr16") %>%
  filter(transcript_length <5000, !is.na(transcript_length)) %>% 
  select(chr, transcript_length)

chr19_lengths <- chr19 %>%
  filter(transcript_type == "mRNA") %>%
  mutate(transcript_length = as.numeric(stop) - as.numeric(start),
         chr = "chr19") %>%
  filter(transcript_length <5000, !is.na(transcript_length)) %>% 
select(chr, transcript_length)

combined_lengths <- bind_rows(chr7_lengths, chr16_lengths, chr19_lengths)
cleaned_lengths <- combined_lengths 
```



#IQR based cleaning for mRNAs  
```{r}
knitr::opts_chunk$set(echo = TRUE)

cleaned_lengths <- combined_lengths %>% 
  group_by(chr) %>%
  mutate(
    Q1 = quantile(transcript_length, 0.25, na.rm = TRUE),
    Q3 = quantile(transcript_length, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    lower_bound = Q1 - 1.5 * IQR,
    upper_bound = Q3 + 1.5 * IQR
  ) %>%
  filter(transcript_length >= lower_bound, transcript_length <= upper_bound) %>%
  select(-Q1, -Q3, -IQR, -lower_bound, -upper_bound) %>%
  ungroup()
```


#boxplot
```{r}
knitr::opts_chunk$set(echo = TRUE)

# Boxplot
ggplot(cleaned_lengths, aes(x = chr, y = transcript_length, fill = chr)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", outlier.shape = 16) +
  labs(title = "Boxplot of Transcript Lengths (IQR-cleaned)",
       x = "Chromosome",
       y = "Transcript Length (bp)") +
  theme_minimal() +
  theme(legend.position = "none")
```

#histogramme
```{r }
knitr::opts_chunk$set(echo = TRUE)

# Histogram + Density (facet ile kromozomlara göre ayrı ayrı)
ggplot(cleaned_lengths, aes(x = transcript_length, fill = chr)) +
  geom_histogram(aes(y = ..density..), bins = 50, alpha = 0.5, position = "identity") +
  geom_density(alpha = 0.6) +
  facet_wrap(~ chr, scales = "free") +
  labs(title = "Histogram and Density of Transcript Lengths (IQR-cleaned)",
       x = "Transcript Length (bp)",
       y = "Density") +
  theme_minimal()
```

#shapiro test to check normal distribution

```{r }
knitr::opts_chunk$set(echo = TRUE)

library(rstatix)
combined_lengths %>%
  group_by(chr) %>% 
shapiro_test(transcript_length) %>% 
#p degerleri coook kucuk o yuzen NORMAL DAGILMIYOR. dolayisiyla NON-PARAMETRIK
view()

```

#kruskal-wallis test 

```{r }
knitr::opts_chunk$set(echo = TRUE)

kruskal.test(transcript_length ~ chr, data = combined_lengths) %>% 
  view()
#gruplar arasinda fark var

```

#dunn's

```{r }
knitr::opts_chunk$set(echo = TRUE)

library(rstatix)

combined_lengths %>%
  dunn_test(transcript_length ~ chr, p.adjust.method = "bonferroni") %>% 
view()
```
